---
title: "Opioids cleanup—redo"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE,
	fig.showtext = TRUE
)
```

```{r}
library(tidyverse)
library(tidycensus)
library(camiller)
library(cwi)
library(lubridate)
```

## Crosswalks
```{r}
big_vill_xwalk <- readxl::read_excel("../raw_data/match_communities_to_towns.xlsx") %>%
  rename(place = Community, town = Town) 

xtra_xwalk <- tribble(
  ~place,                ~town,
  "north grosvenordale", "thompson",
  "new brit",            "new britain",
  "waterbruy",           "waterbury",
  "no haven",            "north haven",
  "w haven",             "west haven",
  "n haven",             "new haven",
  "stafford spgs",       "stafford",
  "vernon-rockville",    "vernon",
  "rockvl",              "vernon",
  "vernon rockvl",       "vernon",
  "sakem",               "salem",
  "s glastonbury",       "glastonbury",
  "north windam",        "windham"
)

op_xwalk <- cwi::xwalk %>%
  distinct(town) %>%
  mutate(place = town) %>%
  bind_rows(
    cwi::village2town %>% select(place, town),
    big_vill_xwalk,
    xtra_xwalk) %>%
  rename(real_town = town) %>%
  mutate_all(tolower) %>%
  distinct(place, real_town) %>%
  arrange(place)

town2county <- tidycensus::fips_codes %>%
  filter(state == "CT") %>%
  distinct(county_code, county) %>%
  mutate(county = str_remove(county, " County")) %>%
  inner_join(
    cwi::xwalk %>% 
      mutate(county_fips = substr(town_fips, 3, 5)) %>%
      distinct(town, county_fips),
    by = c("county_code" = "county_fips")
  ) %>%
  mutate_all(tolower)
```

## Read data
Reading directly from open data portal

```{r}
drugs_read_0 <- RSocrata::read.socrata("https://data.ct.gov/resource/deaths.json") %>% 
  select(date, age, sex, race,
         city = residencecity, county = residencecounty, state = residencestate,
         heroin, oxycodone, methadone, oxymorphone, hydrocodone, fentanyl, tramadol=tramad, morphine_notheroin, opiatenos, fentanylanalogue, hydromorphone, 
         cocaine, ethanol, benzodiazepine, amphet, othersignifican,
         anyopioid, other, 
         cod)
```

As of 05/21/2020, the 2019 data is hosted separately on "https://portal.ct.gov/OCME/Statistics"

Adding 2019 data
```{r}
drugs_2019 <- readxl::read_excel("../raw_data/DrugDeaths_2015-2019.xlsx", sheet="2019") %>% 
  janitor::clean_names() %>% 
  select(date=dod, age, sex, race, city=residence_city, county=residence_county, state=residence_state, heroin, oxycodone, methadone, oxymorphone, hydrocodone, fentanyl, tramadol, morphine_notheroin=morphine_not_heroin, opiatenos=opiate_nos, fentanylanalogue=fentanyl_analogue, hydromorphone, cocaine, ethanol, benzodiazepine=benzodiazepines, amphet=meth_amphetamine, othersignifican=other_significan, anyopioid=any_opioid, other, cod=cause_of_death)

drugs_read <- rbind(drugs_read_0, drugs_2019)
```

## Clean
```{r}
drugs0 <- drugs_read %>%
  as_tibble() %>%
  mutate(age = as.numeric(age)) %>%
  mutate_if(is.character, tolower) %>%
  mutate(city = city %>%
           str_replace_all("\\s{2,}", " ") %>%
           str_trim()) %>%
  mutate_at(vars(sex:state), as.factor) %>%
  mutate(race = race %>%
           fct_collapse(latino = c("hispanic, black", "hispanic, white")) %>%
           fct_collapse(aapi = c("chinese", "korean", "asian indian", 
                                 "asian, other", "hawaiian")) %>% 
           fct_other(keep = c("white", "black", "latino", "aapi"), 
                     other_level = "other_unknown") %>%
           fct_explicit_na(na_level = "other_unknown"),
         sex = fct_explicit_na(sex, na_level = "unknown")) %>%
  mutate(cod = cod %>%
           str_replace_all("[:punct:]", " ") %>%
           str_replace_all("(?<=\\d)\\s", "_") %>%
           str_trim() %>%
           str_replace_all("\\s{2,}", " ") %>%
           str_replace_all("\\bher\\s", "heroin ") %>%
           str_replace_all("\\band\\B", "and ")) %>%
  filter(state == "ct" | is.na(state))
drugs <- drugs0 %>% 
  inner_join(op_xwalk, by = c("city" = "place")) %>%
  select(-county) %>%
  left_join(town2county, by = c("real_town" = "town"))
```

## Text mining
```{r eval=FALSE, include=FALSE}
# most common words that aren't names of drugs
drugs$cod %>% 
  as_tibble() %>% 
  tidytext::unnest_tokens(word, value) %>% 
  anti_join(tidytext::stop_words) %>% 
  count(word, sort=TRUE)
```

```{r}
op_stops <- data.frame(
  word = c("acute", "intoxication", "combined", "effects", "due", "toxicity", "toxicities", "use", "complications", "including", "drug", "multidrug", "multiple", "intoxicationcombined", "chronic", "associated", "recent", "abuse", "using", "substance", "following", "disease", "toxcity"),
  lexicon = "drugs"
) %>%
  bind_rows(tidytext::stop_words)

# to find opioid
drug_strs <- c("\\bfen", "44700", "47700", "br?upre?no", "codeine", "fentanil", "fentanyl", "h-morph", "heroi?n", "hyd-?morph", "hydr-?mor", "hydr?o?morph?", "hydro?\\s?morp", "hydrocodone", "hydrocod", "levorphanol", "meperidine", "methadone", "mor?ph?i+ne", "morph", "morphone", "opiate", "opioi?d", "oxycodone", "oxycod", "oxymorph", "percocet", "poppy", "tapentadol", "tramad")
drug_re <- drug_strs %>%
  paste(collapse = "|") %>%
  sprintf("(%s)", .)

drugs_out <- drugs %>%
  select(date:race, town = real_town, county, heroin:cod) %>%
  gather(key = subs, value, -date:-county, -cod, -other) %>%
  mutate(value = as.factor(value) %>%
           fct_collapse(y = c("y pops", "y (ptch)", "y-a", "yes")) %>%
           fct_other(keep = "y", other_level = "n") %>%
           fct_explicit_na(na_level = "n")) %>%
  filter(value == "y") %>%
  group_by_at(vars(-subs, -value)) %>%
  summarise(subs_txt = c(subs, other, cod) %>%
              na.omit() %>%
              paste(collapse = " ")) %>%
  ungroup() %>%
# unnest tokens to remove duplicate words
  tidytext::unnest_tokens(word, input = subs_txt) %>%
  distinct() %>%
  anti_join(op_stops, by = "word") %>%
  group_by_at(vars(-word)) %>%
  summarise(subs_txt = paste(word, collapse = " ")) %>%
  ungroup() %>%
  mutate(is_opioid = str_detect(subs_txt, drug_re)) %>%
  mutate(age_group = cut(.$age, breaks=c(0,24,34,44,54,64,Inf), 
                         labels=c("14-24", "25-34", "35-44", "45-54", 
                                  "55-64", "65 and over"))) %>% 
  select(date, age, age_group, sex:county, is_opioid, subs_txt)
```

## Write file
```{r}
write_csv(drugs_out, "../output_data/overdoses_indiv_2012_2019.csv")
```

```{r}
drugs %>%
  select(date:race, town = real_town, county, heroin:cod) %>%
  gather(key = subs, value, -date:-county, -cod, -other) %>%
  mutate(value = as.factor(value) %>%
           fct_collapse(y = c("y pops", "y (ptch)", "y-a", "yes")) %>%
           fct_other(keep = "y", other_level = "n") %>%
           fct_explicit_na(na_level = "n")) %>%
  filter(value == "y") %>%
  group_by_at(vars(-subs, -value)) %>%
  summarise(subs_txt = c(subs, other, cod) %>%
              na.omit() %>%
              paste(collapse = " ")) %>%
  ungroup() %>%
# unnest tokens to remove duplicate words
  tidytext::unnest_tokens(word, input = subs_txt) %>%
  distinct() %>%
  anti_join(op_stops, by = "word") %>%
  group_by_at(vars(-word)) %>%
  summarise(subs_txt = paste(word, collapse = " ")) %>%
  ungroup() %>% 
  mutate(subs_txt = str_remove(subs_txt, drug_re))
  
```









## Quick summary stats

Date range: Jan 1, 2012 to Dec 30, 2019
```{r}
range(drugs_out$date)
```

Number of OD deaths (not rate!) has more than triped since 2012.
```{r}
drugs_out %>%
  count(year = year(date))
```

227% percent increase from 2012 to 2019
```{r}
drugs_out %>%
  count(year = year(date)) %>%
  filter(year %in% range(year)) %>%
  spread(key = year, value = n, sep = "_") %>%
  mutate(change = year_2019 - year_2012,
         percent_change = change / year_2012)
```

93 percent of drug related deaths involve opioid/opiate
```{r}
drugs_out %>%
  count(is_opioid) %>%
  mutate(share = n / sum(n))
```

74 percent of drug related deaths are men
```{r}
drugs_out %>%
  filter(!is.na(sex)) %>%
  count(sex) %>%
  mutate(share = n / sum(n))
```

93 percent of men who overdosed involved opioids; 91 percent of women
```{r}
drugs_out %>%
  filter(!is.na(sex)) %>%
  group_by(sex, is_opioid) %>%
  summarise(n = n()) %>%
  mutate(share = n / sum(n))
```

Drug related deaths by race: white 79 percent, Latino 12 percent, Black 8.7 percent, AAPI 0.7 percent.
57 percent are white men.
```{r}
drugs_out %>%
  count(race) %>%
  mutate(share = n / sum(n))
drugs_out %>%
  filter(!is.na(sex)) %>%
  count(race, sex) %>%
  mutate(share = n / sum(n))
```

Age
**how to use age-adjusted #s?**
```{r}
drugs_out %>% count(age_group)
drugs_out %>%
  filter(is_opioid) %>% 
  group_by(age_group) %>% 
  count(year = year(date)) %>%
  mutate(yr = paste("1", "1", year, sep = "-") %>% mdy()) %>% 
  ggplot(aes(x = yr, y = n, color=age_group)) +
    geom_line(size=1.0, se = T) +
    labs(title = "Annual count of opioid overdose deaths by age, 2012-2018",
         subtitle = "Deaths among 25-34 & 35-44 have surpassed 45-54") # how to word?
```

Opioid + other substances
```{r}
# to delete opioids
drug_strs_2 <- c("oxycodone", "oxymorphone",  "morphine", "tramadol", "fentanyl", "hydrocodone", "opiates", "\\bfen", "44700", "47700", "br?upre?no", "codeine", "fentanil", "fentanyl", "h-morph", "heroi?n", "hyd-?morph", "hydr-?mor", "hydr?o?morph?", "hydro?\\s?morp", "hydrocod", "levorphanol", "meperidine", "methadone", "mor?ph?i+ne", "morph", "morphone", "opiate", "opioi?d", "oxycod", "oxymorph", "percocet", "poppy", "tapentadol", "tramad")
drug_re_2 <- drug_strs_2 %>%
  paste(collapse = "|") %>%
  sprintf("(%s)", .)

drugs_out %>% 
  mutate(subs_text_2 = str_remove_all(subs_txt, drug_re_2)) %>% 
  mutate(id = row_number()) %>% 
  tidytext::unnest_tokens(word, input = subs_txt) %>% 
  group_by(id) %>% 
  

```


## Quick plots
```{r}
drugs_out %>%
  count(month = month(date), year = year(date)) %>%
  mutate(mon_yr = paste(month, "1", year, sep = "-") %>% mdy()) %>%
  ggplot(aes(x = mon_yr, y = n)) +
    geom_point() +
    geom_smooth(method = loess, size = 0.4, se = F) +
    labs(title = "Monthly count of overdose deaths, 2012-2018")
```

```{r}
drugs_out %>%
  count(month = month(date), year = year(date), is_opioid) %>%
  mutate(mon_yr = paste(month, "1", year, sep = "-") %>% mdy()) %>%
  ggplot(aes(x = mon_yr, y = n, color = is_opioid)) +
    geom_point() +
    geom_smooth(method = loess, size = 0.4, se = F) +
    labs(title = "Monthly count of overdose deaths, opioid vs non, 2012-2018")
```

```{r}
drugs_out %>%
  filter(!is.na(sex)) %>%
  count(month = month(date), year = year(date), sex) %>%
  mutate(mon_yr = paste(month, "1", year, sep = "-") %>% mdy()) %>%
  ggplot(aes(x = mon_yr, y = n, color = sex)) +
    geom_point() +
    geom_smooth(method = loess, size = 0.4, se = F) +
    labs(title = "Monthly count of overdose deaths by sex, 2012-2019")
```

```{r}
drugs_out %>%
  count(month = month(date), year = year(date), race) %>%
  mutate(mon_yr = paste(month, "1", year, sep = "-") %>% mdy()) %>%
  ggplot(aes(x = mon_yr, y = n, color = race)) +
    geom_point() +
    geom_smooth(method = loess, size = 0.4, se = F) +
    labs(title = "Monthly count of overdose deaths by race, 2012-2019")
```

```{r}
drugs_out %>%
  count(month = month(date), year = year(date)) %>%
  ggplot(aes(x = month, y = n, fill = as.factor(year))) +
    geom_area(position = position_stack()) +
    scale_x_continuous(breaks = 1:12) +
    labs(title = "Monthly count of deaths, 2012-2019")
```







