---
title: "Opioids analysis"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE,
	fig.showtext = TRUE
)
```

```{r}
library(tidyverse)
library(tidycensus)
library(camiller)
library(cwi)
library(lubridate)
```

```{r}
drugs_out <- read_csv("../output_data/overdoses_indiv_2012_2019.csv")
```


Date range: Jan 1, 2012 to Dec 30, 2019
```{r}
range(drugs_out$date)
```

Number of OD deaths (not rate!) has more than triped since 2012.
```{r}
drugs_out %>%
  count(year = year(date))
```

227 percent increase from 2012 to 2019
```{r}
drugs_out %>%
  count(year = year(date)) %>%
  filter(year %in% range(year)) %>%
  spread(key = year, value = n, sep = "_") %>%
  mutate(change = year_2019 - year_2012,
         percent_change = change / year_2012)
```

93 percent of drug related deaths involve opioid/opiate
```{r}
drugs_out %>%
  count(is_opioid) %>%
  mutate(share = n / sum(n))
```

74 percent of drug related deaths are men
```{r}
drugs_out %>%
  filter(!is.na(sex)) %>%
  count(sex) %>%
  mutate(share = n / sum(n))
```

93 percent of men who overdosed involved opioids; 91 percent of women
```{r}
drugs_out %>%
  filter(!is.na(sex)) %>%
  group_by(sex, is_opioid) %>%
  summarise(n = n()) %>%
  mutate(share = n / sum(n))
```

Drug related deaths by race: white 79 percent, Latino 12 percent, Black 8.7 percent, AAPI 0.7 percent.
57 percent are white men.
```{r}
drugs_out %>%
  count(race) %>%
  mutate(share = n / sum(n))
drugs_out %>%
  filter(!is.na(sex)) %>%
  count(race, sex) %>%
  mutate(share = n / sum(n))
```

Age
**how to use age-adjusted #s?**
```{r}
drugs_out %>% count(age_group)
drugs_out %>%
  filter(is_opioid) %>% 
  group_by(age_group) %>% 
  count(year = year(date)) %>%
  mutate(yr = paste("1", "1", year, sep = "-") %>% mdy()) %>% 
  ggplot(aes(x = yr, y = n, color=age_group)) +
    geom_line(size=1.0) +
    labs(title = "Annual opioid overdose deaths by age, 2012-2018",
         subtitle = "Deaths among 25-34 & 35-44 have surpassed 45-54 (wording?)") # how to word?
```

Opioid + other substances
```{r}
# to delete opioids
drug_strs_2 <- c("anyopioid", "oxycodone", "oxymorphone",  "morphine", "tramadol", "fentanyl", "hydrocodone", "opiates", "\\bfen", "44700", "47700", "br?upre?no", "codeine", "fentanil", "fentanyl", "h-morph", "heroi?n", "hyd-?morph", "hydr-?mor", "hydr?o?morph?", "hydro?\\s?morp", "hydrocod", "levorphanol", "meperidine", "methadone", "mor?ph?i+ne", "morph", "morphone", "opiate", "opioi?d", "oxycod", "oxymorph", "percocet", "poppy", "tapentadol", "tramad")
drug_re_2 <- drug_strs_2 %>%
  paste(collapse = "|") %>%
  sprintf("(%s)", .)

# top 5 most common
drugs_out %>% 
  mutate(subs_txt_2 = str_remove_all(subs_txt, drug_re_2)) %>% 
  mutate(id = row_number()) %>% 
  tidytext::unnest_tokens(word, input = subs_txt_2) %>% 
  count(word) %>% arrange(desc(n))
# 1537

drugs_out_top5_2019 <- drugs_out %>% 
  mutate(benzodiazepine = str_detect(subs_txt, "benzodiazepine|benzodiazepines|benzo$|benzodiazapine")) %>% 
  mutate(cocaine = str_detect(subs_txt, "cocain|cocanie")) %>% 
  mutate(ethanol = str_detect(subs_txt, "ethanol|ethanl")) %>% 
  mutate(alprazolam = str_detect(subs_txt, "alprazolam")) %>% 
  mutate(alcohol = str_detect(subs_txt, "alcohol")) %>% 
  filter(is_opioid==TRUE) %>% 
  filter(date > "2018-12-31 18:00:00")

drugs_out_top5_2019 %>% 
  select(benzodiazepine:alcohol) %>% 
  gather(key, value) %>%
  filter(value) %>% 
  ggplot(aes(key)) +
  geom_bar(stat="count", position = "dodge") # add alprazolam: brand name = xanax

all5 <- drugs_out_top5_2019 %>% 
  summarize(any_true = sum( benzodiazepine|cocaine|ethanol|alprazolam|alcohol) / 1011)

benzo_alc <- drugs_out_top5_2019 %>% 
  summarize(any_true = sum( benzodiazepine|alcohol) / 1011)

coc_xan_eth <- drugs_out_top5_2019 %>% 
  summarize(any_true = sum( cocaine|ethanol|alprazolam) / 1011)
```
62 percent of opioid related deaths in 2019 also involved cocaine, ethanol, and/or alprazolam.

Because opioid overdoses kill people by suppressing their ability to breathe, many opioid-involved od deaths also include other drugs that interact on this respiratory depression effect. In 2019, 33% of opioid related deaths also involved benzodiazepines and/or alcohol.

Heroin vs. synthetic
```{r}
semi <- "heroin|codone"
synth <- "fentanyl|tramadol|methadone"

drugs_out_3 <- drugs_out %>% 
  filter(is_opioid) %>% 
  mutate(semi = str_detect(subs_txt, semi)) %>% 
  mutate(synth = str_detect(subs_txt, synth))

semi <- drugs_out_3 %>% 
  count(year = year(date), semi, synth) %>% 
  group_by(semi, year) %>% 
  summarize(semi_count = sum(n)) %>% 
  filter(semi) %>% 
  ungroup()

synth <- drugs_out_3 %>% 
  count(year = year(date), semi, synth) %>% 
  group_by(synth, year) %>% 
  summarize(synth_count = sum(n)) %>% 
  filter(synth) %>% 
  ungroup()

full_join(semi, synth, by="year") %>% 
  select(year, semi_count, synth_count) %>% 
  pivot_longer(cols=semi_count:synth_count) %>% 
  ggplot(aes(x=year, y=value, color=name)) +
  geom_line()
```

## Fentanyl
```{r}
drugs_out_4 <- drugs_out %>% 
  mutate(is_fentanyl = str_detect(subs_txt, "fentanyl|fentayl")) %>% 
  count(year = year(date), is_fentanyl) %>% 
  group_by(is_fentanyl, year) %>% 
  summarize(fentanyl_count = sum(n)) %>% 
  ungroup()

drugs_out_4 %>% ggplot(aes(x=year, y=fentanyl_count, color=is_fentanyl)) +
  geom_line()
```
In the past 5 years, fentanyl involvement in drug overdose deaths has increased by 380%.

## Quick plots
```{r}
drugs_out %>%
  count(month = month(date), year = year(date)) %>%
  mutate(mon_yr = paste(month, "1", year, sep = "-") %>% mdy()) %>%
  ggplot(aes(x = mon_yr, y = n)) +
    geom_point() +
    geom_smooth(method = loess, size = 0.4, se = F) +
    labs(title = "Monthly count of overdose deaths, 2012-2018")
```

```{r}
drugs_out %>%
  count(month = month(date), year = year(date), is_opioid) %>%
  mutate(mon_yr = paste(month, "1", year, sep = "-") %>% mdy()) %>%
  ggplot(aes(x = mon_yr, y = n, color = is_opioid)) +
    geom_point() +
    geom_smooth(method = loess, size = 0.4, se = F) +
    labs(title = "Monthly count of overdose deaths, opioid vs non, 2012-2018")
```

```{r}
drugs_out %>%
  filter(!is.na(sex)) %>%
  count(month = month(date), year = year(date), sex) %>%
  mutate(mon_yr = paste(month, "1", year, sep = "-") %>% mdy()) %>%
  ggplot(aes(x = mon_yr, y = n, color = sex)) +
    geom_point() +
    geom_smooth(method = loess, size = 0.4, se = F) +
    labs(title = "Monthly count of overdose deaths by sex, 2012-2019")
```

```{r}
drugs_out %>%
  count(month = month(date), year = year(date), race) %>%
  mutate(mon_yr = paste(month, "1", year, sep = "-") %>% mdy()) %>%
  ggplot(aes(x = mon_yr, y = n, color = race)) +
    geom_point() +
    geom_smooth(method = loess, size = 0.4, se = F) +
    labs(title = "Monthly count of overdose deaths by race, 2012-2019")
```

```{r}
drugs_out %>%
  count(month = month(date), year = year(date)) %>%
  ggplot(aes(x = month, y = n, fill = as.factor(year))) +
    geom_area(position = position_stack()) +
    scale_x_continuous(breaks = 1:12) +
    labs(title = "Monthly count of deaths, 2012-2019")
```







